---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---
### Loading and preprocessing the data
```{r readData}
df<- read.csv(unz("repdata-data-activity.zip", "activity.csv"), header=TRUE ,sep=",")
df$DateField<-as.Date(df$date)
```

### What is mean total number of steps taken per day?
```{r createPlotSums}
library(lattice)
totalStepsByDate<-aggregate(df$steps, by=list(df$date),FUN=sum)
names(totalStepsByDate)<-c("Date","Steps")
meanPerDay<-mean(totalStepsByDate$Steps,na.rm = TRUE)
medianPerDay<-median(totalStepsByDate$Steps,na.rm = TRUE)
totalStepsPerDayPlot<-barchart(Steps ~ Date, data=totalStepsByDate,
                               main="Total Number of Steps taken Daily",
                               ylab="Total number of steps", xlab="Date", 
                               scales = list(y=list(tick.number=6), 
                                             x=list(cex=0.6,tick.number=7,rot=80)
                                             ),panel=function(x,y,...){
                               panel.barchart(x,y,...)
                               panel.text(x=x[15],y=21000,paste(c("  Mean Steps per day: ",format(meanPerDay,digits=5)),sep="",collapse=" "))
                               panel.text(x=x[15],y=19500,paste(c("Median Steps per day: ",format(medianPerDay,digits=5)),sep="",collapse=" "))
                               })
print(totalStepsPerDayPlot)
```

**Figure 1.** *Histogram of total number of steps taken in a day for a single individual over the period of study. Number steps in 5 minute intervals were collected for each day over 61 days.*

```{r printFig1, echo=FALSE, results='hide'}
trellis.device(device = "png", filename="figures/Plot1-TotalStepsByDate.png",height=400, width=700)
print(totalStepsPerDayPlot)
dev.off()
```
The daily **mean** total number of the steps taken per day is **`r as.integer(meanPerDay)`** and the **median** total number of steps taken per day is **`r as.integer(medianPerDay)`**.  *Missing values were excluded in computing the mean and median.*

### What is the average daily activity pattern?
```{r createPlotMeans}
library(lattice)
avgActivity<-aggregate(df$steps, by=list(df$interval),FUN=mean, na.rm=TRUE)
names(avgActivity)<-c("Interval","Activity")
yMax.y<-max(avgActivity$Activity)
yMax.x<-avgActivity$Interval[which.max(avgActivity$Activity)]
yMax.time<-strftime( as.POSIXct( "1970-01-01" ) + as.difftime( yMax.x, units="mins" ), "%I:%M %p" )
maxIntervalLabel<-paste(c("Maximum Activity Interval: ",yMax.x," (",yMax.time,")"),sep="", collapse="")
maxActivityLabel<-paste(c("         Maximum Activity: ",format(yMax.y,digits=3)," Steps"),sep="", collapse="")
avgActivityOver5Mins<-xyplot(Activity ~ Interval, data=avgActivity, type="l", 
                             main="Average activity over 5 minute intervals\n (Across all days)", 
                             xlab="Minutes", ylab="Average Activity (Number of Steps)",
                             panel=function(x,y,...){
                               panel.xyplot(x,y,...)
                               panel.abline(v =yMax.x,col = "red", lwd = 1, lty="dashed")
                               panel.text(x=yMax.x+800,y=yMax.y,labels=maxIntervalLabel)
                               panel.text(x=yMax.x+800,y=yMax.y-15,labels=maxActivityLabel)
                               })
print(avgActivityOver5Mins)
````
````{r printFig2, echo=FALSE, results='hide'}
trellis.device(device = "png", filename="figures/Plot2-AverageActivityOver5MinIntervals.png",height=400, width=700)
print(avgActivityOver5Mins)
dev.off()
```

**Figure 2.** *Time series plot of the average number of steps taken during 5-minute intervals across all days. The plot show higher activity during the 800 to 930 minutes (1:20 PM to 3:15 PM), possibly due to vigorous exercise. (b) The 5-minute interval corresponding to `r yMax.x` minutes (around `r yMax.time`) had an average maximum of `r format(yMax.y,digits=3)` steps.*

### Imputing missing values
```{r imputingNas }
dfi<- read.csv(unz("repdata-data-activity.zip", "activity.csv"), header=TRUE ,sep=",")
dfi$DateField<-as.Date(dfi$date)
dfi$stepsImp<-dfi$steps

dateStr<-unique(dfi$DateField)
for (i in 1:length(dateStr)){
  a<-dfi$stepsImp[dfi$DateField==dateStr[i]]
  #atleast one NA and 
  if(any(is.na(a))){
    #has atleast one value
    if(any(!is.na(a))){
      a<-dfi$stepsImp[dfi$DateField==dateStr[i]]
      amean<-mean(a,na.rm=TRUE)
      a[is.na(a)==TRUE]=amean
      dfi$stepsImp[dfi$DateField==dateStr[i]]=a
      }else{
        # has all NAs ==> replace with 0
        dfi$stepsImp[dfi$DateField==dateStr[i]]=0
        }
    }
  }
```

There are `r sum(is.na(dfi$steps))` NAs out of total of `r length(dfi$steps)` observations. Since NAs affect the summary statstics, we explore its variation upon replacing NAs. Current plot shows the affect of replacing the NA values with value for daily mean number of steps for that day. If NAs are observed for an entire day, the mean is assumed to be 0, and all NAs are replaced by 0.
```{r avgDailyImputedVsNormal}
avgsAll<-aggregate(dfi$steps, by=list(dfi$interval),FUN=mean, na.rm=TRUE)
avgImputed<-aggregate(dfi$stepsImp, by=list(dfi$interval),FUN=mean, na.rm=TRUE)
avgsAll$imp<-avgImputed$x
names(avgsAll)<-c("Interval","Steps","Steps-MeanImputed")

totalStepsByDate1<-aggregate(dfi$stepsImp, by=list(dfi$date),FUN=sum)
names(totalStepsByDate1)<-c("Date","StepsImp")

meanPerDay1<-mean(totalStepsByDate1$StepsImp,na.rm = TRUE)
medianPerDay1<-median(totalStepsByDate1$StepsImp,na.rm = TRUE)

library(ggplot2)
library(reshape2)
avgsAll2<-melt(data=avgsAll, id.vars="Interval")
avgImpPlot<-ggplot(data = avgsAll2, aes(x = Interval, y = value, colour = variable)) + geom_line()
avgImpPlot<-avgImpPlot  + labs(x = "Minutes") + labs(y = "Number of steps") + labs(title = "Effect of imputing NAs\n with mean for the day")

print(avgImpPlot)
```
````{r printFig3, echo=FALSE, results='hide'}
trellis.device(device = "png", filename="figures/Plot3-AverageActivityAffectOfImupting.png",height=400, width=700)
print(avgImpPlot)
dev.off()
```
**Figure 3.** *Time series plot of the average number of steps taken during 5-minute intervals across all days (a) NAs not imputed mean = `r format(meanPerDay,digits=3)` and median = `r format(medianPerDay,digits=3)` (b) Imputing NAs with mean for the day changes the statists to mean = `r format(meanPerDay1,digits=3)` and median = `r format(medianPerDay1,digits=3)`. If all values for a day are NAs, then mean is assumed to be 0.*

**Imputing with NAs with mean lowers the daily Mean and Median number os steps taken.**

### Are there differences in activity patterns between weekdays and weekends?

```{r weekdayOrWeekend}
library(lattice)
df$DateField<-as.Date(df$date)
df$DayOfWeek<-factor(weekdays(df$DateField))
df$Weekday<-factor(weekdays(df$DateField))
levels(df$Weekday)<-list(`weekday`=c("Monday","Tuesday","Wednesday","Thursday","Friday"), `weekend`=c("Saturday","Sunday"))
avgStepsByWeekday<-aggregate(df$steps, by=list(df$interval,df$Weekday),FUN=mean, na.rm=TRUE)
names(avgStepsByWeekday)<-c("Interval","Weekday","MeanSteps")
weekdaysTable<-table(df$Weekday)/288
weekdayPlot<-xyplot(MeanSteps ~ Interval|Weekday, data=avgStepsByWeekday, nrow=2,ncol=1, type="l", 
                    main="Weekday or Weekend\n Average activity over 5 minute intervals\n (Across all days)", 
                    xlab="Interval", ylab="Number of Steps")

print(weekdayPlot)
````
````{r printFig4, echo=FALSE, results='hide'}
trellis.device(device = "png", filename="figures/Plot4-WeekDyVsWeekendActivity.png",height=700, width=400)
print(weekdayPlot)
dev.off()
```
**Figure 4.** *Average activity pattern during weekdays and weekends. The plot show definitive higher acivity during the afternoon period, corresponding to about 800-950 minutes during weekdays(`r weekdaysTable[["weekday"]]`), while during weekends(`r weekdaysTable[["weekend"]]`) the activity was higher throughout the day. This is possibly due to a vigorous exercise schedule during weekdays that is no apparent during weekend.*

```{r hideRefs, echo=FALSE, results='hide'}
### see: http://www.stat.columbia.edu/~gelman/arm/missing.pdf
### random.imp <- function (a){
### missing <- is.na(a)
### n.missing <- sum(missing)
### a.obs <- a[!missing]
##### imputed <- a
##### imputed[missing] <- sample (a.obs, n.missing, replace=TRUE)
##### return (imputed)
##### }
###### (To see how this function works, take a small dataset and evaluate the function line
###### by line.) We use random.imp to create a completed data vector of earnings:
###### R code earnings.imp <- random.imp (earnings)
###### visualizing missing values marginplot(
###### http://cran.r-project.org/web/packages/mice/mice.pdf
###### http://teachpress.environmentalinformatics-marburg.de/2013/07/creating-publication-quality-graphs-in-r-7/

```