---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---
### Loading and preprocessing the data
```{r readData}
df<- read.csv(unz("repdata-data-activity.zip", "activity.csv"), header=TRUE ,sep=",")
df$DateField<-as.Date(df$date)
```

### What is mean total number of steps taken per day?
```{r createPlotSums}
library(lattice)
totalStepsByDate<-aggregate(df$steps, by=list(df$date),FUN=sum)
names(totalStepsByDate)<-c("Date","Steps")
meanPerDay<-mean(totalStepsByDate$Steps,na.rm = TRUE)
medianPerDay<-median(totalStepsByDate$Steps,na.rm = TRUE)
totalStepsPerDayPlot<-barchart(Steps ~ Date, data=totalStepsByDate,
                               main="Total Number of Steps taken Daily",
                               ylab="Total number of steps", xlab="Date", 
                               scales = list(y=list(tick.number=6), 
                                             x=list(cex=0.6,tick.number=7,rot=80)
                                             ),panel=function(x,y,...){
                               panel.barchart(x,y,...)
                               panel.text(x=x[15],y=21000,paste(c("  Mean Steps per day: ",format(meanPerDay,digits=5)),sep="",collapse=" "))
                               panel.text(x=x[15],y=19500,paste(c("Median Steps per day: ",format(medianPerDay,digits=5)),sep="",collapse=" "))
                               })
print(totalStepsPerDayPlot)
```

**Figure 1.** *Histogram of total number of steps taken in a day for a single individual over the period of study. Number steps in 5 minute intervals were collected for each day over 61 days.*

```{r printFig1, echo=FALSE, results='hide'}
trellis.device(device = "png", filename="figures/Plot1-TotalStepsByDate.png",height=400, width=700)
print(totalStepsPerDayPlot)
dev.off()
```
The daily **mean** total number of the steps taken per day is **`r as.integer(meanPerDay)`** and the **median** total number of steps taken per day is **`r as.integer(medianPerDay)`**.  *Missing values were excluded in computing the mean and median.*

### What is the average daily activity pattern?
```{r createPlotMeans}
library(lattice)
avgActivity<-aggregate(df$steps, by=list(df$interval),FUN=mean, na.rm=TRUE)
names(avgActivity)<-c("Interval","Activity")
yMax.y<-max(avgActivity$Activity)
yMax.x<-avgActivity$Interval[which.max(avgActivity$Activity)]
yMax.time<-strftime( as.POSIXct( "1970-01-01" ) + as.difftime( yMax.x, units="mins" ), "%I:%M %p" )
maxIntervalLabel<-paste(c("Maximum Activity Interval: ",yMax.x," (",yMax.time,")"),sep="", collapse="")
maxActivityLabel<-paste(c("         Maximum Activity: ",format(yMax.y,digits=3)," Steps"),sep="", collapse="")
avgActivityOver5Mins<-xyplot(Activity ~ Interval, data=avgActivity, type="l", 
                             main="Average activity over 5 minute intervals\n (Across all days)", 
                             xlab="Minutes", ylab="Average Activity (Number of Steps)",
                             panel=function(x,y,...){
                               panel.xyplot(x,y,...)
                               panel.abline(v =yMax.x,col = "red", lwd = 1, lty="dashed")
                               panel.text(x=yMax.x+800,y=yMax.y,labels=maxIntervalLabel)
                               panel.text(x=yMax.x+800,y=yMax.y-15,labels=maxActivityLabel)
                               })
print(avgActivityOver5Mins)
````
````{r printFig2, echo=FALSE, results='hide'}
trellis.device(device = "png", filename="figures/Plot2-AverageActivityOver5MinIntervals.png",height=400, width=700)
print(avgActivityOver5Mins)
dev.off()
```

**Figure 2.** *Time series plot of the average number of steps taken during 5-minute intervals across all days. The plot show higher activity during the 800 to 930 minutes (1:20 PM to 3:15 PM), possibly due to vigorous exercise. (b) The 5-minute interval corresponding to `r yMax.x` minutes (around `r yMax.time`) had an average maximum of `r format(yMax.y,digits=3)` steps.*

### Imputing missing values
```{r}

```





### Are there differences in activity patterns between weekdays and weekends?

```{r weekdayOrWeekend}
library(lattice)
df$DateField<-as.Date(df$date)
df$DayOfWeek<-factor(weekdays(df$DateField))
df$Weekday<-factor(weekdays(df$DateField))
levels(df$Weekday)<-list(`weekday`=c("Monday","Tuesday","Wednesday","Thursday","Friday"), `weekend`=c("Saturday","Sunday"))
avgStepsByWeekday<-aggregate(df$steps, by=list(df$interval,df$Weekday),FUN=mean, na.rm=TRUE)
names(avgStepsByWeekday)<-c("Interval","Weekday","MeanSteps")
weekdaysTable<-table(df$Weekday)/288
weekdayPlot<-xyplot(MeanSteps ~ Interval|Weekday, data=avgStepsByWeekday, nrow=2,ncol=1, type="l", 
                    main="Weekday or Weekend\n Average activity over 5 minute intervals\n (Across all days)", 
                    xlab="Interval", ylab="Number of Steps")

print(weekdayPlot)
````
````{r printFig4, echo=FALSE, results='hide'}
trellis.device(device = "png", filename="figures/Plot4-WeekDyVsWeekendActivity.png",height=700, width=400)
print(weekdayPlot)
dev.off()
```
**Figure 4.** *Average activity pattern during weekdays and weekends. The plot show definitive higher acivity during the afternoon period, corresponding to about 800-950 minutes during weekdays(`r weekdaysTable[["weekday"]]`), while during weekends(`r weekdaysTable[["weekend"]]`) the activity was higher throughout the day. This is possibly due to a vigorous exercise schedule during weekdays that is no apparent during weekend.*

```{r hideRefs, echo=FALSE, results='hide'}
### see: http://www.stat.columbia.edu/~gelman/arm/missing.pdf
### random.imp <- function (a){
### missing <- is.na(a)
### n.missing <- sum(missing)
### a.obs <- a[!missing]
##### imputed <- a
##### imputed[missing] <- sample (a.obs, n.missing, replace=TRUE)
##### return (imputed)
##### }
###### (To see how this function works, take a small dataset and evaluate the function line
###### by line.) We use random.imp to create a completed data vector of earnings:
###### R code earnings.imp <- random.imp (earnings)
###### visualizing missing values marginplot(
###### http://cran.r-project.org/web/packages/mice/mice.pdf
###### http://teachpress.environmentalinformatics-marburg.de/2013/07/creating-publication-quality-graphs-in-r-7/

```